---
title: "STA 101 - HW 1"
author: "Mingyi Xue"
date: "April 5, 2018"
output: html_document
---

### Load databases into current workingspace:

```{r, echo = FALSE, warning=FALSE}
hospital <- read.csv("~/STA 101/Homework/HW1/hospital.csv", header = TRUE)
head(hospital, n=3)
poverty <- read.csv("~/STA 101/Homework/HW1/poverty.csv", header = TRUE)
head(poverty, n=3)
rehab <- read.csv("~/STA 101/Homework/HW1/rehab.csv", header = TRUE)
head(rehab, n=3)
```

### Load package `ggplot2`:

```{r}
library(ggplot2)
```

### Problem 1
**(a)**
```{r, echo = FALSE, warning=FALSE}
qplot(PovPct, Brth15to17,data = poverty)+
  ggtitle("Scatter plot of PovPCt and Birth15to17")+
  xlab("Poverty percent")+
  ylab("Birth Rate")
```

The plot suggests a linear relationship between PovPct and Brth15to17.

**(b)**
```{r, echo = FALSE, warning=FALSE}
the.correlation = round(cor(poverty$Brth15to17, poverty$PovPct),4)
```

The estimated correlation between PovPct and Brth15to17 is `r the.correlation` and this suggests a strong linear relationship.

**(c)**
```{r, echo = FALSE, warning=FALSE}
the.five = fivenum(poverty$PovPct)
the.five
```
minimun:`r the.five[1]`  
the first quartile:`r the.five[2]`  
median:`r the.five[3]`  
the third quartile:`r the.five[4]`  
maximum:`r the.five[5]`  

**(d)**
```{r, echo = FALSE, warning=FALSE}
ggplot(poverty,aes(y = Brth15to17,x = factor("")))+
  geom_boxplot()+
  ggtitle("Boxplot for Birth rate")+
  ylab("Birth rate")+ 
  xlab("")+
  coord_flip()
```

The default criteria to set outliers in R is:  
upper_boundary=the third quartile + (the third quartile - the first quartile) * 1.5  
lower_boundary=the first quartile - (the third quartile - the first quartile) * 1.5  
According to the boxplot of Brth15to17, there is an exremely large value outlier.  

**(e)**
```{r, echo = FALSE, warning=FALSE}
the.mean = mean(poverty$PovPct)
the.sd = sd(poverty$PovPct)
the.left_boundary = round(the.mean - 3 * the.sd, 4)
the.right_boundary = round(the.mean + 3 * the.sd, 4)
```
Use the formula:  
the.left_boundary = mean - 3 * sd  
the.right_boundary = mean + 3 * sd  
An appropriate range for PovPct is from `r the.left_boundary` to `r the.right_boundary`.

### Problem 2
**(a)**
```{r, echo = FALSE, warning=FALSE}
the.line = lm(Brth15to17~PovPct, data = poverty)
the.coef = round(the.line$coefficients,4)
the.coef
```

The estimated linear regression line of Brth15to17 by PovPct is  
$\hat{y}=`r the.coef[1]`+ `r the.coef[2]`X_{1}$

**(b)**
```{r, echo = FALSE, warning=FALSE}

```

The estimated slope is $\hat{\beta_{1}} = `r the.coef[2]`$. It means when the percent of the states population living in poverty increases by 1 percent, the estimated value of birth rate per 1000 females 15 to 17 years old will increase by `r the.coef[2]` percent on average.

**(c)**
```{r, echo = FALSE, warning=FALSE}

```

The estimated intercept is $\hat{\beta_{0}} = `r the.coef[1]`$. It is inapproriate because a state cannot be absolutely poverty-free in reality.

**(d)**
```{r, echo = FALSE, warning=FALSE}
y_i = the.coef[1] + the.coef[2] * 10
```

The estimated average value of Brth15to17 is `r y_i` at PovPct equals 10.

**(e)**
```{r, echo = FALSE, warning=FALSE}
the.summary = summary(the.line)
```

The value of $R^{2}$ is `r round(the.summary$r.squared, 3)`.  
It means if we use linear regression model to estimate the value of Brth15to17 instead of the average value, the total error can be reduced by $53.3\%$.

### Problem 3
**(a)**
```{r, echo = FALSE, warning=FALSE}
ggplot(rehab, aes(x = fitness, y = days))+
  geom_boxplot()+
  xlab("fitness")+
  ylab("days")+
  ggtitle("Grouped boxplot of days by fitness level")+
  coord_flip()
```

The grouped boxplot shows that there is a great difference between the groups because of little overlap.

**(b)**
```{r, echo = FALSE, warning=FALSE}
ggplot(rehab, aes(x = factor(""), y = days))+
  geom_boxplot()+
  xlab("")+
  ylab("days")+
  ggtitle("Boxplot of days")+
  coord_flip()
```

No outliers in $days$.

**(c)**
```{r, echo = FALSE, warning=FALSE}

```

There will be 3 $\beta$'s coreesponding to $fitness$.  
$\beta_{0}$:the estimated average value of number of days for category $AboveAve$ to complete the therapy.  
$\beta_{1}$:the estimated average difference of number of days to comlete the therapy between $Ave$ and $AboveAve$.   
$\beta_{2}$:the estimated average difference of number of days to comlete the therapy between $BelowAve$ and $Ave$.  

Corresponding $X_{i}$'s are X_Ave and X_Bel.

**(d)**
```{r, echo = FALSE, warning=FALSE}
the.five = fivenum(rehab$days)
```

minimun:`r the.five[1]`  
the first quartile:`r the.five[2]`  
median:`r the.five[3]`  
the third quartile:`r the.five[4]`  
maximum:`r the.five[5]`  

**(e)**
```{r, echo = FALSE, warning=FALSE}

```

Yes. Although $days$ have no outliers in its boxplot as a whole, there are outliers when categorized by $fitness$.

### Problem 4
**(a)**
```{r, echo = FALSE, warning=FALSE}
the.line=lm(days ~ fitness, data = rehab)
the.coef = round(the.line$coefficients,4)
```
```{r,warning = FALSE}
unique(rehab$fitness)
```

$\hat{y} = `r the.coef[1]` + `r the.coef[2]`X_{Ave} + `r the.coef[3]`X_{Bel}$ 

**(b)**
```{r, echo = FALSE, warning=FALSE}

```

$\hat{\beta}_{1}$ means the estimated average value of number of days that people in category Ave need to complete the therapy is `r the.coef[2]` more than people in category AboveAve.

**(c)**
```{r, echo = FALSE, warning=FALSE}

```

$\hat{\beta}_{2}$ means the estimated average value of number of days that people in category BelowAve need to complete the therapy is  `r the.coef[3]` more than people in category AboveAve.

**(d)**
```{r, echo = FALSE, warning=FALSE}

```

$\hat{\beta}_{0}$ means the estimated average value of number of days that people in category AboveAve need to complete the therapy is `r the.coef[1]`.

**(e)**
```{r, echo = FALSE, eval = FALSE}

```
category fitness=AboveAve:$\hat{y}=`r the.coef[1]`$  
category fitness=Ave:$\hat{y}=`r the.coef[1]+the.coef[2]`$  
category fitness=BelowAve:$\hat{y}=`r the.coef[1]+the.coef[3]`$

### Problem 5
**(a)**
```{r, echo = FALSE, warning=FALSE}
qplot(Stay, InfctRsk, data = hospital)+
  ggtitle("Scatter plot of InfctRsk by Stay")+
  xlab("Stay")+
  ylab("InfcRsk")

the.correlation1 = round(cor(hospital$InfctRsk, hospital$Stay),4)
```

There are several outliers on hospital dataset. Correlation between InfctRsk and Stay is `r the.correlation1`, which indicates a medium linear relationship between this two random variables.

**(b)**
```{r, echo = FALSE, warning=FALSE}
ggplot(hospital, aes(x = MedSchool, y = InfctRsk))+geom_boxplot()+
  ggtitle("Boxplot of InfctRsk grouped by MedSchool")+
  xlab("MedSchool")+
  ylab("InfcRsk")+
  coord_flip()
```

There is no apparent difference between different categories of InfctRsk because two boxplots overlap a lot. 

**(c)**
```{r, echo = FALSE, warning=FALSE}
the.correlation1 = cor(hospital$InfctRsk, hospital$Stay)
```

The correlation between InfctRsk and Stay is `r round(the.correlation1,4)`.
MedSchool is not a numeric variable.

**(d)**
```{r, echo = FALSE, warning=FALSE}

```

Yes. There are outliers on the boxplot of InfctRsk grouped by MedSchool.

**(e)**
```{r, echo = FALSE, warning=FALSE}
the.mean=mean(hospital$Stay)
the.sd=sd(hospital$Stay)
```

A reasonable range of average patient stay is from `r round(the.mean-3*the.sd,3)` to `r round(the.mean+3*the.sd,3)`.

### Problem 6
**(a)**
```{r, echo = FALSE, warning=FALSE}
the.line=lm(InfctRsk ~ ., data = hospital)
the.coef = round(the.line$coefficients,4)
the.line
```

The regression equation is: $\hat{y} = `r the.coef[1]` + `r the.coef[2]`X_{Yes}+ `r the.coef[3]`X_1$.

**(b)**
```{r, echo = FALSE, warning=FALSE}
the.temp=the.coef[1] + the.coef[2]
```

hospitals not in a teaching school: $\hat{y} = `r the.coef[1]` + `r the.coef[3]`X_1$  
hospitals in a teaching school:$\hat{y} = `r the.temp` + `r the.coef[3]`X_1$

**(c)**
```{r, echo = FALSE, warning=FALSE}

```

$\hat{\beta}_{1}$ means the percentage of patients who get a secondary infection during their hospital stay will be `r the.coef[2]` lager if they are in teaching school associated with the hospital than those who are not.

**(d)**
```{r, echo = FALSE, warning=FALSE}

```

$\hat{\beta}_{2}$ means the percentage of patients who get a secondary infection during their hospital stay will increase by `r the.coef[3]` on average if the average length of their stay increase by 1 day.


**(e)**
```{r, echo = FALSE, warning=FALSE}
x.star =data.frame(Stay=7, MedSchool= "No")
the.result=predict(the.line, x.star)
```

The estimated value of infection percentage is `r round(the.result,4)` on average.

### Problem 7
**(a)**
```{r, echo = FALSE, warning=FALSE}
the.line=lm(InfctRsk~MedSchool+Stay+MedSchool*Stay, data=hospital)
the.coef=round(the.line$coefficients, 4)
the.line

```
The regression model with an interaction term is $\hat{y}=`r the.coef[1]`+`r the.coef[2]`X_{1}+`r the.coef[3]`X_{Yes}`r the.coef[4]`X_{1}X_{Yes}$.

**(b)**
```{r, echo = FALSE, warning=FALSE}

```
category MedSchool=Yes:$\hat{y}=`r the.coef[1]+the.coef[3]`+`r the.coef[2]+the.coef[4]`X_{1}$  
category MedSchool=No:$\hat{y}=`r the.coef[1]`+`r the.coef[2]`X_{1}$

**(c)**
```{r, echo = FALSE, warning=FALSE}

```

The slope of category MedSchool=Yes means the estimated value of InfctRsk will increase by `r the.coef[2]+the.coef[4]` on average when Stay increases by 1 day, holding categorical variable MedSchool equals Yes.  
The slope of category MedSchool=No means the estimated value of InfctRsk will increase by `r the.coef[2]` on average when Stay increases by 1 day, holding categorical variable MedSchool equals No.

**(d)**
```{r, echo = FALSE, warning=FALSE}

```

The intercept of category MedSchool=Yes means the estimated average value of InfctRsk is `r the.coef[1]+the.coef[3]` when Stay equals 0.  
The intercept of category MedSchool=No means the estimated average value of InfctRsk is `r the.coef[1]` when Stay equals 0.
These two intercept is inappropriate because a patient's stay in hospital could not be 0.

**(e)**
```{r, echo = FALSE, warning=FALSE}
x.star=data.frame(MedSchool="No",Stay=7)
y_i=round(predict(the.line, x.star),4)
```

The estimated value is `r y_i` on average at MedSchool="No" and Stay=7.

### Problem 8
**(a)**
```{r, echo = FALSE, warning=FALSE}

```

False. Correlation only measures the strength of linear relationship.


**(b)**
```{r, echo = FALSE, warning=FALSE}

```
False. A regression model with interation terms can be presented as $\hat{y}=\hat{\beta}_{0}+\hat{\beta}_{1}X_{1}+\hat{\beta}_{2}X_{2}+\hat{\beta}_{3}X_{1}X_{2}$, it is the same with the model of non-interaction terms if and only if $\hat{\beta}_{3}=0$, which scarcely happens.

**(c)**
```{r, echo = FALSE, warning=FALSE}

```

True. $r$ range from $-1$ to $1$.  
If $r>0$ and $r<=1$, this suggests a positive linear relationship. Else if $r<0$ and $r>=-1$,this suggests a negative linear relationship. Else if $r=0$, this suggests no linear relationship.

**(d)**
```{r, echo = FALSE, warning=FALSE}

```

True. $e_{i} = y_{i} - \hat{y}_{i}$ when $e_{i} > 0$, it means $y_{i} > \hat{y}_{i}$, which indicates that we underestimated the value.


### R Appendix
```{r, ref.label=knitr::all_labels(),echo=TRUE,eval=FALSE}
```